---
title: 'EN.553.632. Bayesian Statistics -- Final Project'
author: "Zhiwei Gong, Xu He, Bingxu Han"
output:
  html_document: default
  pdf_document: default
date: "2022-12-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require('ggplot2')) 
{
  install.packages('ggplot2');
  library(ggplot2);
}

if (!require('reshape')) 
{
  install.packages('reshape');
  library(reshape);
}

if (!require('MASS')) 
{
  install.packages('MASS');
  library(MASS);
}

if (!require('coda')) 
{
  install.packages('coda');
  library(coda);
}

if (!require('car')) 
{
  install.packages('car');
  library(car);
}

if (!require('psych')) 
{
  install.packages('psych');
  library(psych);
}

if (!require('rcompanion')) 
{
  install.packages('rcompanion');
  library(rcompanion);
}

if (!require('dplyr')) 
{
  install.packages('dplyr');
  library(dplyr);
}

if (!require('tidyr')) 
{
  install.packages('tidyr');
  library(tidyr);
}

if (!require('corrgram')) 
{
  install.packages('corrgram');
  library(corrgram);
}

if (!require('corrplot')) 
{
  install.packages('corrplot');
  library(corrplot);
}

```


### Introduction

### Dataset
```{r}
# read dataset
insurance = read.csv('./data/insurance.csv')
insurance
```
## Data analysis
```{r}
# Data analysis
# check for missing values
f = function(x) {
  any(is.na(x))
}
check.na = apply(insurance,2,f)
check.na

# Transform variables to factor
insurance$sex = factor(insurance$sex)
insurance$smoker = factor(insurance$smoker)
insurance$region = factor(insurance$region)

# Display structure of dataset
str(insurance)
```

```{r}
df = insurance
df <- as.data.frame(lapply(df,as.numeric))
df
```

```{r}
corrgram(df,
         cex.labels = 0.8,         
         lower.panel="panel.pts",   
         upper.panel="panel.cor",   
         diag.panel="panel.density",
         cor.method="pearson")      

```
```{r}
ggplot(insurance, aes(x = bmi, y = charges, colour = smoker)) +  geom_point()
ggplot(insurance, aes(x = children, y = charges, colour = smoker)) +  geom_point()
```


```{r}
table(insurance$sex)
```
```{r}
type <- c('Male','Female')
nums <- c(676,662)
p_df <- data.frame(sex = type, nums = nums)
label_value <- paste('(', round(p_df$nums/sum(p_df$nums) * 100, 1), '%)', sep = '')
label <- paste(p_df$type, label_value, sep = '')
p <- ggplot(data = p_df, mapping = aes(x = 'Content', y = nums, fill = sex)) + geom_bar(stat = 'identity', position = 'stack', width = 1) + coord_polar(theta = 'y')+  coord_polar(theta = 'y') + labs(x = '', y = '', title = '') + theme(axis.text = element_blank()) + theme(axis.ticks = element_blank())+ scale_fill_discrete(labels = label) + theme(panel.grid=element_blank())
p
```
```{r}
table(insurance$children)
type <- c('0','1','2','3','4','5')
nums <- c(574,324, 240, 157 , 25,  18)
p_df <- data.frame(children = type, nums = nums)
label_value <- paste('(', round(p_df$nums/sum(p_df$nums) * 100, 1), '%)', sep = '')
label <- paste(p_df$type, label_value, sep = '')
p <- ggplot(data = p_df, mapping = aes(x = 'Content', y = nums, fill = children)) + geom_bar(stat = 'identity', position = 'stack', width = 1) + coord_polar(theta = 'y')+  coord_polar(theta = 'y') + labs(x = '', y = '', title = '') + theme(axis.text = element_blank()) + theme(axis.ticks = element_blank())+ scale_fill_discrete(labels = label)+ theme(panel.grid=element_blank())
p
```
```{r}
table(insurance$smoker)
type <- c('no','yes')
nums <- c(1064 , 274)
p_df <- data.frame(smoker = type, nums = nums)
label_value <- paste('(', round(p_df$nums/sum(p_df$nums) * 100, 1), '%)', sep = '')
label <- paste(p_df$type, label_value, sep = '')
p <- ggplot(data = p_df, mapping = aes(x = 'Content', y = nums, fill = smoker)) + geom_bar(stat = 'identity', position = 'stack', width = 1) + coord_polar(theta = 'y')+  coord_polar(theta = 'y') + labs(x = '', y = '', title = '') + theme(axis.text = element_blank()) + theme(axis.ticks = element_blank())+ scale_fill_discrete(labels = label)+ theme(panel.grid=element_blank())
p
```
```{r}
table(insurance$region)
type <- c('northeast', 'northwest', 'southeast', 'southwest')
nums <- c(324   ,    325    ,   364   ,    325)
p_df <- data.frame(region = type, nums = nums)
label_value <- paste('(', round(p_df$nums/sum(p_df$nums) * 100, 1), '%)', sep = '')
label <- paste(p_df$type, label_value, sep = '')
p <- ggplot(data = p_df, mapping = aes(x = 'Content', y = nums, fill = region)) + geom_bar(stat = 'identity', position = 'stack', width = 1) + coord_polar(theta = 'y')+  coord_polar(theta = 'y') + labs(x = '', y = '', title = '') + theme(axis.text = element_blank()) + theme(axis.ticks = element_blank())+ scale_fill_discrete(labels = label)+ theme(panel.grid=element_blank())
p

```


```{r}
# summary of whole dataset
summary(insurance)

# Plot the distribution of insurance response variable "Charges"
par(mfrow = c(1,2)) # combine the two plots
hist(insurance$charges, main = "Histogram of charges", col = "lightblue")
plot(density(insurance$charges), main = "Density plot of charges")
polygon(density(insurance$charges), col = "orange")
```
In regression situations, highly skewed data can result in a pooly fitting model. When all of the values are positive, the log (natural logarithm) transformation that can often be used to normalize highly skewed data.
```{r}
insurance$charges <- log(insurance$charges)
summary(insurance$charges)
par(mfrow = c(1,2)) 
hist(insurance$charges, main = "Histogram of charges", col = "lightblue")
plot(density(insurance$charges), main = "Density plot of charges")
polygon(density(insurance$charges), col = "orange")
```
We can see this log transformation gives us a much less skewed and more symmetric distrubution.\

## Check the outliers using the boxplot.
```{r}
par(mfrow = c(1,3))
boxplot(insurance$age, main = "Histogram of age")
boxplot(insurance$bmi, main = "Histogram of bmi")
boxplot(insurance$children, main = "Histogram of children")
```

We can see the 'bmi' has outliters. Remove the outliers from the data set.
```{r}
outliers_remover <- function(a){
  df <- a
  aa <- c()
  count <- 1
  for(i in 1:ncol(df)){
    if(is.numeric(df[,i])){
      Q3 <- quantile(df[,i], 0.75, na.rm = TRUE)
      Q1 <- quantile(df[,i], 0.25, na.rm = TRUE) 
      IQR <- Q3 - Q1  #IQR(df[,i])
      upper <- Q3 + 1.5 * IQR
      lower <- Q1 - 1.5 * IQR
      for(j in 1:nrow(df)){
        if(is.na(df[j,i]) == TRUE){
            next
        }
            
        else if(df[j,i] > upper | df[j,i] < lower){
          aa[count] <- j
          count <- count+1                  
        }
      }
    }
  }
  
  df <- df[-aa,]
}

insurance_new <- outliers_remover(insurance)
str(insurance_new)
```

Compare original data, we can see the observations reduced from 1338 to 1329.\

## Model Building
```{r}
# stepwise regression
model1 = lm(formula = insurance_new$charges ~ insurance_new$age  + insurance_new$smoker + insurance_new$bmi + insurance_new$sex + insurance_new$children, data = insurance_new)
summary(model1)

model2 = lm(formula = insurance_new$charges ~ insurance_new$age  + insurance_new$smoker + insurance_new$bmi + insurance_new$sex, data = insurance_new)
summary(model2)

model3 = lm(formula = insurance_new$charges ~ insurance_new$age  + insurance_new$smoker + insurance_new$bmi, data = insurance_new)
summary(model3)
```
```{r}
#Check the BIC and AIC. 
#Sort the result for BIC and AIC values with model numbers.
com_mod <- compareLM(model1, model2, model3)
com_model <- com_mod$Fit.criteria
com_model[order(com_model$AIC),]
```

```{r}
# Perform Diagnostic
plot(model1)

hist(residuals(model1))
```

```{r}
# Improve Model Performance 
# new variables
insurance_new$age2 <- insurance_new$age^2
insurance_new$bmi30 <- ifelse(insurance_new$bmi >= 30, 1, 0)
insurance_new$bmi2 <- insurance_new$bmi^2
```

```{r}
# new model
Improved_model1 = lm(formula = insurance_new$charges ~ insurance_new$age  + insurance_new$bmi + insurance_new$smoker + insurance_new$sex + insurance_new$children + insurance_new$age2 + insurance_new$bmi2 + (insurance_new$smoker * insurance_new$bmi30), data = insurance_new)
summary(Improved_model1)

```
```{r}
# Compare the performance of original model 1 and improved model 1 - correlation accuracy
insurance = read.csv('./data/insurance.csv')
insurance$charges <- log(insurance$charges)
insurance_new <- outliers_remover(insurance)

# Create Training and Test data -- model1 (before the improvments)
trainingRowIndex <- sample(1:nrow(insurance_new), 0.7 * nrow(insurance_new)) #row indices for training data
trainingData <- insurance_new[trainingRowIndex,] #training data
testData <- insurance_new[-trainingRowIndex,] #test data
# Build the model on training data
model1 <- lm(charges ~ age + smoker + bmi + sex + children, data = trainingData)
# predict
pred1 <- predict(model1, testData)
actuals_preds1 <- data.frame(cbind(actuals = testData$charges, predicted = pred1))
cor(actuals_preds1)

```

```{r}
insurance = read.csv('./data/insurance.csv')
insurance$charges <- log(insurance$charges)
insurance_new <- outliers_remover(insurance)
insurance_new$age2 <- insurance_new$age^2
insurance_new$bmi2 <- insurance_new$bmi^2
insurance_new$bmi30 <- ifelse(insurance_new$bmi >= 30, 1, 0)

# Create Training and Test data -- model2 (after the improvments)
trainingRowIndex <- sample(1:nrow(insurance_new), 0.7 * nrow(insurance_new)) #row indices for training data
trainingData <- insurance_new[trainingRowIndex,] #training data
testData <- insurance_new[-trainingRowIndex,] #test data
# Build the model on training data
model2 <- lm(charges ~ age + smoker + bmi + sex + children + age2 + bmi2 + smoker*bmi30, data = trainingData)
# predict
pred2 <- predict(model2, testData)
actuals_preds2 <- data.frame(cbind(actuals = testData$charges, predicted = pred2))
cor(actuals_preds2) # correlation accuracy
mape <- mean(abs((actuals_preds2$predicted - actuals_preds2$actuals))/actuals_preds2$actuals)  # mean absolute percentage deviation
mape
res = actuals_preds2$actuals - actuals_preds2$predicted
rmse = sqrt(mean(res^2))
rmse
```

### MCMC

```{r}
# Bayesian analysis of OLS linear regression models
# Bayesian analysis: specify priors for the unknown regression coefficients beta and variance sigma^2
insurance = read.csv('./data/insurance.csv')
insurance$charges <- log(insurance$charges)
insurance_new <- outliers_remover(insurance)
y = insurance_new$charges
n = length(y)
insurance_new$smoker<-ifelse(insurance_new$smoker=="yes",1,0)
X = cbind(rep(1, n), insurance_new$age, insurance_new$smoker, insurance_new$bmi)
p = dim(X)[2]
beta.ols = solve(t(X)%*%X)%*%t(X)%*%y
beta.ols

fit.ls = lm(charges ~ age + smoker + bmi, data = insurance_new)
beta.0 = rep(0, p)
Sigma.0 = diag(c(197, 0.16, 35))
nu.0 = 1
sigma2.0 = 15^2

beta.0<-fit.ls$coef
nu.0<-1  ; sigma2.0<-sum(fit.ls$res^2)/(n-p)
Sigma.0<- solve(t(X)%*%X)*sigma2.0*n

S<-5000
####

rmvnorm<-function(n,mu,Sigma) 
{ # samples from the multivariate normal distribution
  E<-matrix(rnorm(n*length(mu)),n,length(mu))
  t(  t(E%*%chol(Sigma)) +c(mu))
}
###

### some convenient quantites
n<-length(y)
p<-length(beta.0)
iSigma.0<-solve(Sigma.0)
XtX<-t(X)%*%X

### store mcmc samples in these objects
B = 500
beta.post<-matrix(nrow=S+B,ncol=p)
sigma2.post<-rep(NA,S+B)

### starting value
set.seed(1)
sigma2<- var( residuals(lm(y~0+X)) )
### MCMC algorithm
for( scan in 1:(S+B)) {

#update beta
V.beta<- solve(  iSigma.0 + XtX/sigma2 )
E.beta<- V.beta%*%( iSigma.0%*%beta.0 + t(X)%*%y/sigma2 )
beta<-t(rmvnorm(1, E.beta,V.beta) )

#update sigma2
nu.n<- nu.0+n
ss.n<-nu.0*sigma2.0 + sum(  (y-X%*%beta)^2 )
sigma2<-1/rgamma(1,nu.n/2, ss.n/2)

#save results of this scan
beta.post[scan,]<-beta
sigma2.post[scan]<-sigma2
                        }

#####
round( apply(beta.post,2,mean), 3)
Autocorrelation_beta0 = acf(beta.post[,1][1 : (S+B)], main = 'ACF for beta0', plot = FALSE)
Autocorrelation_beta1 = acf(beta.post[,2][1 : (S+B)], main = 'ACF for beta1', plot = FALSE)
Autocorrelation_beta2 = acf(beta.post[,3][1 : (S+B)], main = 'ACF for beta2', plot = FALSE)
Autocorrelation_beta3 = acf(beta.post[,4][1 : (S+B)], main = 'ACF for beta3', plot = FALSE)
```

```{r}
insurance = read.csv('./data/insurance.csv')
insurance$charges <- log(insurance$charges)
insurance_new <- outliers_remover(insurance)

# Create Training and Test data -- model2 (after the improvments)
trainingRowIndex <- sample(1:nrow(insurance_new), 0.7 * nrow(insurance_new)) #row indices for training data
trainingData <- insurance_new[trainingRowIndex,] #training data
testData <- insurance_new[-trainingRowIndex,] #test data
# Build the model on training data
beta0 = rep(7.059, 399)
beta1 = rep(0.035, 399)
beta2 = rep(1.538, 399)
beta3 = rep(0.011, 399)
testData$smoker <- ifelse(testData$smoker =='yes', 1, 0)
pred2 = beta0 + beta1 * testData$age + beta2 * testData$smoker + beta3 * testData$bmi
# predict
#pred2 <- predict(model2, testData)
actuals_preds2 <- data.frame(cbind(actuals = testData$charges, predicted = pred2))
cor(actuals_preds2) # correlation accuracy
mape <- mean(abs((actuals_preds2$predicted - actuals_preds2$actuals))/actuals_preds2$actuals)  # mean absolute percentage deviation
mape
res = actuals_preds2$actuals - actuals_preds2$predicted
rmse = sqrt(mean(res^2))
rmse
```


```{r}
round( mean(sigma2.post), 3)
par(mfrow=c(3,2))
acf(beta.post[,1][1 : (S+B)], main = 'ACF for beta0')
```
```{r}
acf(beta.post[,2][1 : (S+B)], main = 'ACF for beta1')
```

```{r}
acf(beta.post[,3][1 : (S+B)], main = 'ACF for beta2')
```

```{r}
acf(beta.post[,4][1 : (S+B)], main = 'ACF for beta3')
```

```{r}
acf(sigma2.post[1 : (S+B)], main = 'ACF for sigma^2')
```


```{r}
Autocorrelation_beta0
s_mu_tau.df = data.frame(beta.post)
colnames(s_mu_tau.df) = c('beta0', 'beta1', 'beta2', 'beta3')
s_mu_tau.df$sigma2 = sigma2.post
s_mu_tau.df$s = 1:(S+B)
cut_size = 10
s_mu_tau.df = s_mu_tau.df %>%
  tibble %>%
  mutate(scut = cut(s, breaks = cut_size)) %>%
  gather('variable', 'value', beta0:sigma2)
ggplot(s_mu_tau.df, aes(x = scut, y = value)) +
  facet_wrap(~ variable, scales = 'free_y') +
  geom_boxplot() +
  theme(axis.text.x = element_blank()) +
  xlab('Samples')
```






